noam_beta_diversity

### 2024-09-03 noam_beta_diversity.Rmd
Compleed site level dissimilarity over time and site level ordination for all algae. working on macro now

```{r import and format data}
# starting with all data now because of 0 algae transects/quadrats
data <- read.csv(here::here("data", "allbenthicdata_cleaned_colnames.csv"), stringsAsFactors = F)

@@

str(data)

meta_cols <- c(colnames(data[1:7] ), "island_side", "island_side_habitat", "stratum")
species_cols <- colnames(data)[!colnames(data) %in% meta_cols]

meta_df <- data[ , which(names(data) %in% meta_cols)]

site_colours <- c("lter_1" = "cornflowerblue",
                  "lter_2" = "darkblue",
                  "lter_3" = "yellowgreen",
                  "lter_4" = "forestgreen",
                  "lter_5" = "mediumorchid",
                  "lter_6" = "orange")
island_side_colours <- c("north" = "darkblue",
                         "east" = "mediumorchid",
                         "west" = "orange")
habitat_linetypes <- c("fringing" = "solid",
                       "backreef" = "longdash",
                       "outer_10" = "twodash",
                       "outer_17" = "dotted")
habitat_shapes <- c("fringing" = 16,
                     "backreef" = 17,
                     "outer_10" = 1,
                     "outer_17" = 2)

#pivot long 
# long_data <- pivot_longer(
#   data = data,
#   cols = 8:77,
#   names_to = "taxa",
#   values_to = "percent_cover") 
# 
# 
#  transect_level <- long_data %>%
#    dplyr::group_by(year, island_side, site, habitat, transect, taxa) %>%
#    dplyr::summarise(mean_cover = mean(percent_cover), 
#              SE = sd(percent_cover)/sqrt(n()))
#  
# site_level <- long_data %>%
#   dplyr::group_by(year, island_side, site, habitat, site_habitat, taxa) %>%
#   dplyr::summarise(mean_site_cover = mean(percent_cover), 
#             site_se = sd(percent_cover)/sqrt(n()))
# 
# subset based on habitat
fringing <- subset(data, habitat == "fringing")
backreef <- subset(data, habitat == "backreef")
outer10 <- subset(data, habitat == "outer_10")
outer17 <- subset(data, habitat == "outer_17")

@@

```{r fringing bray curtis}
fringing_species_matrix <- as.matrix(fringing[ , -which(names(fringing) %in% meta_cols)])
fringing_species_matrix <- fringing_species_matrix/100 # convert from percent to proportion
rownames(fringing_species_matrix) <- paste(fringing$year,"-", fringing$location, sep = "")

>>>>

```{r import and format data}
# for figures
island_side_colours <- c("north" = "darkblue",
                         "east" = "mediumorchid",
                         "west" = "orange")
habitat_colours <- c("fringing" = "#D81B60",
                     "backreef" = "#FFC107",
                     "outer_10" = "#1E88E5",
                     "outer_17" = "#004D40")

# some quadrats have dissimilarity = 0 or 1
# Smithson verkuilen transformation for beta-regression
sv_trans <- function(prop, N, s = 0.000005){
  (prop*(nrow(N) - 1) + s)/nrow(N)
  # where prop is the proportional value you're transforming, 
  # N is the sample size, which is specified by taking the number of rows/observations from a given dataframe,
  # and s is a small offset 
}

# starting with all data now because of 0 algae transects/quadrats
data <- read.csv(here::here("data", "allbenthicdata_cleaned_colnames.csv"), stringsAsFactors = F)

@@

macroalgae <- data %>% 
  select(-algal_turf, -crustose_corallines, -cyanophyta, -damselfish_turf, -lithophyllum_kotschyanum,
         -neogoniolithon_brassica_florida, -sporolithon_sp)

sand_coral <- read_csv(here("data", "allbenthicdata_cleaned_colnames.csv")) %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, sand, coral)

# get ranges to calculate over (this is just for indexing purposes in the next bit of code)
startAlg <- grep("acanthophora_spicifera", colnames(macroalgae))
endAlg <- grep("valonia_ventricosa", colnames(macroalgae))

macroalgae %>% 
  # bring in sand data
  full_join(sand_coral) %>% 
  rename(coral_cover = coral,
         sand_cover = sand) %>% 
  mutate(hard_cover = 100-sand_cover) %>% 
  mutate(macroalgae_cover = rowSums(macroalgae[startAlg:endAlg]), # percent cover of macroalgae

         macroalgae_cover_hard = case_when(
           sand_cover == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (macroalgae_cover/hard_cover)*100 
           ),
         # also for corals
         coral_cover_hard = case_when(
           sand_cover == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (coral_cover/hard_cover)*100 
           )
         ) %>% # percent cover of hard substrate
   # make a year-habitat column for betadisper
  mutate(hab_year = paste0(habitat, "_", year),
         site_hab_year = paste0(site, "_", hab_year)) %>% 
  select(year, location, site, habitat, island_side, site_habitat, island_side_habitat,
         hab_year, site_hab_year, transect, quadrat, macroalgae_cover, 
         macroalgae_cover_hard, sand_cover, hard_cover, coral_cover, coral_cover_hard)  -> meta_df

meta_df$all_algae_cover <- rowSums(data[startAlg:endAlg]) # percent cover of all algae
meta_df$all_algae_cover_hard <- case_when(
           meta_df$sand_cover == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (meta_df$all_algae_cover/meta_df$hard_cover)*100   
         )

data <- merge(meta_df, data )
macroalgae <- merge(meta_df, macroalgae)

meta_cols <- colnames(meta_df)

full_species_cols <- colnames(data)[!colnames(data) %in% meta_cols]
macro_species_cols <- colnames(macroalgae)[!colnames(macroalgae) %in% meta_cols]

# summarize at site level
site_full <- data %>%
  dplyr::group_by(year, island_side, site, habitat, site_habitat, hab_year, site_hab_year) %>%
  dplyr::summarize(across(all_of(
    c("macroalgae_cover", "macroalgae_cover_hard", "sand_cover", "hard_cover", "coral_cover",
      "coral_cover_hard","all_algae_cover", "all_algae_cover_hard", full_species_cols)), mean, na.rm = TRUE))

# summarize at species level
site_macro <- macroalgae %>%
  dplyr::group_by(year, island_side, site, habitat, site_habitat, hab_year, site_hab_year) %>%
  dplyr::summarize(across(all_of(
    c("macroalgae_cover", "macroalgae_cover_hard", "sand_cover", "hard_cover", "coral_cover",
      "coral_cover_hard","all_algae_cover", "all_algae_cover_hard", macro_species_cols)), mean, na.rm = TRUE))

# extract site level meta_df for future merges
site_meta_df <- site_full[,which(names(site_full) %in% meta_cols)]

fringing <- subset(data, habitat == "fringing")
backreef <- subset(data, habitat == "backreef")
outer10 <- subset(data, habitat == "outer_10")
outer17 <- subset(data, habitat == "outer_17")

fringing_macro <- subset(macroalgae, habitat == "fringing")
backreef_macro <- subset(macroalgae, habitat == "backreef")
outer10_macro <- subset(macroalgae, habitat == "outer_10")
outer17_macro <- subset(macroalgae, habitat == "outer_17")

fringing_site <- subset(site_full, habitat == "fringing")
backreef_site <- subset(site_full, habitat == "backreef")
outer10_site <- subset(site_full, habitat == "outer_10")
outer17_site <- subset(site_full, habitat == "outer_17")

fringing_site_macro <- subset(site_macro, habitat == "fringing")
backreef_site_macro <- subset(site_macro, habitat == "backreef")
outer10_site_macro <- subset(site_macro, habitat == "outer_10")
outer17_site_macro <- subset(site_macro, habitat == "outer_17")
```

Moving all site level analyses to before quadrat for beta diversity

```{r site level bray curtis setup}
fringing_site_species_matrix <- as.matrix(fringing_site[ , -which(names(fringing_site) %in% meta_cols)])
fringing_site_species_matrix <- fringing_site_species_matrix/fringing_site$hard_cover # convert from percent to proportion of hard cover
rownames(fringing_site_species_matrix) <- paste(fringing_site$year,"-", fringing_site$site_habitat, sep = "")

# filter rare species
# fringing_site_species_presence <- colSums(fringing_site_species_matrix > 0) # Presence absence matrix
# fringing_site_species_percentage <- (fringing_site_species_presence/nrow(fringing_site_species_matrix)) # proportion of rows where a species is present
# fringing_site_common_species <- names(fringing_site_species_percentage[fringing_site_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
# fringing_site_species_matrix <- fringing_site_species_matrix[,colnames(fringing_site_species_matrix) %in% fringing_site_common_species] # filter original species matrix

fringing_site_rows_to_remove <- rowSums(fringing_site_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
fringing_site_species_matrix <- fringing_site_species_matrix[!fringing_site_rows_to_remove,]

fringing_site_species_matrix_untransformed <- fringing_site_species_matrix
fringing_site_species_matrix_transformed <- asin(sqrt(fringing_site_species_matrix)) # arcsin squareroot transform
fringing_site_subset <- fringing_site[!fringing_site_rows_to_remove,] # make sure rows are also removed from dataframe
fringing_site_subset$stratum <- paste(fringing_site_subset$year,"-", fringing_site_subset$site_habitat, sep = "") # for merging with bray matrix later

brayfringing_site <- vegdist(fringing_site_species_matrix_transformed, "bray") # create bray dissimilarity matrix
# View(brayfringing_site)
brayfringing_site_untransformed <- vegdist(fringing_site_species_matrix_untransformed, "bray")
fringing_site_braymat <- as.matrix(brayfringing_site) # conver to matrix object
# convert to pairwise comparison
fringing_site_comp <- fringing_site_braymat[lower.tri(fringing_site_braymat)] # extract lower triangle
fringing_site_row_indices <- row(fringing_site_braymat)[lower.tri(fringing_site_braymat)] # extract row indices for first quad in comparison
fringing_site_col_indices <- col(fringing_site_braymat)[lower.tri(fringing_site_braymat)] # extract col indices = second quad in comp
fringing_site_loc1 <- rownames(fringing_site_braymat)[fringing_site_row_indices] # populate appropriate vector
fringing_site_loc2 <- colnames(fringing_site_braymat)[fringing_site_col_indices] # populate appropriate vector
fringing_site_pairwise_df <- data.frame(loc1 = fringing_site_loc1, loc2 = fringing_site_loc2, dissimilarity = fringing_site_comp)
fringing_site_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", fringing_site_pairwise_df$loc1)) # extract timepoint from first quad in comp
fringing_site_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", fringing_site_pairwise_df$loc2)) # repeat for second quad
fringing_site_pairwise_df$time <- abs(fringing_site_pairwise_df$year2 - fringing_site_pairwise_df$year1) # extract length of differences
fringing_site_pairwise_df$site_1 <- sub(".*-(.*)", "\\1", fringing_site_pairwise_df$loc1) # extract quadrat id from first quad in comp
fringing_site_pairwise_df$site_2 <- sub(".*-(.*)", "\\1", fringing_site_pairwise_df$loc2) # repeat for second quad

fringing_site_untransformed_braymat <- as.matrix(brayfringing_site_untransformed) # conver to matrix object
# convert to pairwise comparison
fringing_site_untransformed_comp <- fringing_site_untransformed_braymat[lower.tri(fringing_site_untransformed_braymat)] # extract lower triangle
fringing_site_untransformed_row_indices <- row(fringing_site_untransformed_braymat)[lower.tri(fringing_site_untransformed_braymat)] # extract row indices for first quad in comparison
fringing_site_untransformed_col_indices <- col(fringing_site_untransformed_braymat)[lower.tri(fringing_site_untransformed_braymat)] # extract col indices = second quad in comp
fringing_site_untransformed_loc1 <- rownames(fringing_site_untransformed_braymat)[fringing_site_untransformed_row_indices] # populate appropriate vector
fringing_site_untransformed_loc2 <- colnames(fringing_site_untransformed_braymat)[fringing_site_untransformed_col_indices] # populate appropriate vector
fringing_site_untransformed_pairwise_df <- data.frame(loc1 = fringing_site_untransformed_loc1, loc2 = fringing_site_untransformed_loc2, dissimilarity = fringing_site_untransformed_comp)
fringing_site_untransformed_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", fringing_site_untransformed_pairwise_df$loc1)) # extract timepoint from first quad in comp
fringing_site_untransformed_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", fringing_site_untransformed_pairwise_df$loc2)) # repeat for second quad
fringing_site_untransformed_pairwise_df$time <- abs(fringing_site_untransformed_pairwise_df$year2 - fringing_site_untransformed_pairwise_df$year1) # extract length of differences
fringing_site_untransformed_pairwise_df$site_1 <- sub(".*-(.*)", "\\1", fringing_site_untransformed_pairwise_df$loc1) # extract quadrat id from first quad in comp
fringing_site_untransformed_pairwise_df$site_2 <- sub(".*-(.*)", "\\1", fringing_site_untransformed_pairwise_df$loc2) # repeat for second quad

# backreef

backreef_site_species_matrix <- as.matrix(backreef_site[ , -which(names(backreef_site) %in% meta_cols)])
backreef_site_species_matrix <- backreef_site_species_matrix/backreef_site$hard_cover # convert from percent to proportion of hard cover
rownames(backreef_site_species_matrix) <- paste(backreef_site$year,"-", backreef_site$site_habitat, sep = "")

# filter rare species
# backreef_site_species_presence <- colSums(backreef_site_species_matrix > 0) # Presence absence matrix
# backreef_site_species_percentage <- (backreef_site_species_presence/nrow(backreef_site_species_matrix)) # proportion of rows where a species is present
# backreef_site_common_species <- names(backreef_site_species_percentage[backreef_site_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
# backreef_site_species_matrix <- backreef_site_species_matrix[,colnames(backreef_site_species_matrix) %in% backreef_site_common_species] # filter original species matrix

backreef_site_rows_to_remove <- rowSums(backreef_site_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
backreef_site_species_matrix <- backreef_site_species_matrix[!backreef_site_rows_to_remove,]

backreef_site_species_matrix_untransformed <- backreef_site_species_matrix
backreef_site_species_matrix_transformed <- asin(sqrt(backreef_site_species_matrix)) # arcsin squareroot transform
backreef_site_subset <- backreef_site[!backreef_site_rows_to_remove,] # make sure rows are also removed from dataframe
backreef_site_subset$stratum <- paste(backreef_site_subset$year,"-", backreef_site_subset$site_habitat, sep = "") # for merging with bray matrix later

braybackreef_site <- vegdist(backreef_site_species_matrix_transformed, "bray") # create bray dissimilarity matrix
# View(braybackreef_site)
braybackreef_site_untransformed <- vegdist(backreef_site_species_matrix_untransformed, "bray")
backreef_site_braymat <- as.matrix(braybackreef_site) # conver to matrix object
# convert to pairwise comparison
backreef_site_comp <- backreef_site_braymat[lower.tri(backreef_site_braymat)] # extract lower triangle
backreef_site_row_indices <- row(backreef_site_braymat)[lower.tri(backreef_site_braymat)] # extract row indices for first quad in comparison
backreef_site_col_indices <- col(backreef_site_braymat)[lower.tri(backreef_site_braymat)] # extract col indices = second quad in comp
backreef_site_loc1 <- rownames(backreef_site_braymat)[backreef_site_row_indices] # populate appropriate vector
backreef_site_loc2 <- colnames(backreef_site_braymat)[backreef_site_col_indices] # populate appropriate vector
backreef_site_pairwise_df <- data.frame(loc1 = backreef_site_loc1, loc2 = backreef_site_loc2, dissimilarity = backreef_site_comp)
backreef_site_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", backreef_site_pairwise_df$loc1)) # extract timepoint from first quad in comp
backreef_site_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", backreef_site_pairwise_df$loc2)) # repeat for second quad
backreef_site_pairwise_df$time <- abs(backreef_site_pairwise_df$year2 - backreef_site_pairwise_df$year1) # extract length of differences
backreef_site_pairwise_df$site_1 <- sub(".*-(.*)", "\\1", backreef_site_pairwise_df$loc1) # extract quadrat id from first quad in comp
backreef_site_pairwise_df$site_2 <- sub(".*-(.*)", "\\1", backreef_site_pairwise_df$loc2) # repeat for second quad

backreef_site_untransformed_braymat <- as.matrix(braybackreef_site_untransformed) # conver to matrix object
# convert to pairwise comparison
backreef_site_untransformed_comp <- backreef_site_untransformed_braymat[lower.tri(backreef_site_untransformed_braymat)] # extract lower triangle
backreef_site_untransformed_row_indices <- row(backreef_site_untransformed_braymat)[lower.tri(backreef_site_untransformed_braymat)] # extract row indices for first quad in comparison
backreef_site_untransformed_col_indices <- col(backreef_site_untransformed_braymat)[lower.tri(backreef_site_untransformed_braymat)] # extract col indices = second quad in comp
backreef_site_untransformed_loc1 <- rownames(backreef_site_untransformed_braymat)[backreef_site_untransformed_row_indices] # populate appropriate vector
backreef_site_untransformed_loc2 <- colnames(backreef_site_untransformed_braymat)[backreef_site_untransformed_col_indices] # populate appropriate vector
backreef_site_untransformed_pairwise_df <- data.frame(loc1 = backreef_site_untransformed_loc1, loc2 = backreef_site_untransformed_loc2, dissimilarity = backreef_site_untransformed_comp)
backreef_site_untransformed_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", backreef_site_untransformed_pairwise_df$loc1)) # extract timepoint from first quad in comp
backreef_site_untransformed_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", backreef_site_untransformed_pairwise_df$loc2)) # repeat for second quad
backreef_site_untransformed_pairwise_df$time <- abs(backreef_site_untransformed_pairwise_df$year2 - backreef_site_untransformed_pairwise_df$year1) # extract length of differences
backreef_site_untransformed_pairwise_df$site_1 <- sub(".*-(.*)", "\\1", backreef_site_untransformed_pairwise_df$loc1) # extract quadrat id from first quad in comp
backreef_site_untransformed_pairwise_df$site_2 <- sub(".*-(.*)", "\\1", backreef_site_untransformed_pairwise_df$loc2) # repeat for second quad

# outer10

outer10_site_species_matrix <- as.matrix(outer10_site[ , -which(names(outer10_site) %in% meta_cols)])
outer10_site_species_matrix <- outer10_site_species_matrix/outer10_site$hard_cover # convert from percent to proportion of hard cover
rownames(outer10_site_species_matrix) <- paste(outer10_site$year,"-", outer10_site$site_habitat, sep = "")

# filter rare species
# outer10_site_species_presence <- colSums(outer10_site_species_matrix > 0) # Presence absence matrix
# outer10_site_species_percentage <- (outer10_site_species_presence/nrow(outer10_site_species_matrix)) # proportion of rows where a species is present
# outer10_site_common_species <- names(outer10_site_species_percentage[outer10_site_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
# outer10_site_species_matrix <- outer10_site_species_matrix[,colnames(outer10_site_species_matrix) %in% outer10_site_common_species] # filter original species matrix

outer10_site_rows_to_remove <- rowSums(outer10_site_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
outer10_site_species_matrix <- outer10_site_species_matrix[!outer10_site_rows_to_remove,]

outer10_site_species_matrix_untransformed <- outer10_site_species_matrix
outer10_site_species_matrix_transformed <- asin(sqrt(outer10_site_species_matrix)) # arcsin squareroot transform
outer10_site_subset <- outer10_site[!outer10_site_rows_to_remove,] # make sure rows are also removed from dataframe
outer10_site_subset$stratum <- paste(outer10_site_subset$year,"-", outer10_site_subset$site_habitat, sep = "") # for merging with bray matrix later

brayouter10_site <- vegdist(outer10_site_species_matrix_transformed, "bray") # create bray dissimilarity matrix
# View(brayouter10_site)
brayouter10_site_untransformed <- vegdist(outer10_site_species_matrix_untransformed, "bray")
outer10_site_braymat <- as.matrix(brayouter10_site) # conver to matrix object
# convert to pairwise comparison
outer10_site_comp <- outer10_site_braymat[lower.tri(outer10_site_braymat)] # extract lower triangle
outer10_site_row_indices <- row(outer10_site_braymat)[lower.tri(outer10_site_braymat)] # extract row indices for first quad in comparison
outer10_site_col_indices <- col(outer10_site_braymat)[lower.tri(outer10_site_braymat)] # extract col indices = second quad in comp
outer10_site_loc1 <- rownames(outer10_site_braymat)[outer10_site_row_indices] # populate appropriate vector
outer10_site_loc2 <- colnames(outer10_site_braymat)[outer10_site_col_indices] # populate appropriate vector
outer10_site_pairwise_df <- data.frame(loc1 = outer10_site_loc1, loc2 = outer10_site_loc2, dissimilarity = outer10_site_comp)
outer10_site_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", outer10_site_pairwise_df$loc1)) # extract timepoint from first quad in comp
outer10_site_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", outer10_site_pairwise_df$loc2)) # repeat for second quad
outer10_site_pairwise_df$time <- abs(outer10_site_pairwise_df$year2 - outer10_site_pairwise_df$year1) # extract length of differences
outer10_site_pairwise_df$site_1 <- sub(".*-(.*)", "\\1", outer10_site_pairwise_df$loc1) # extract quadrat id from first quad in comp
outer10_site_pairwise_df$site_2 <- sub(".*-(.*)", "\\1", outer10_site_pairwise_df$loc2) # repeat for second quad

outer10_site_untransformed_braymat <- as.matrix(brayouter10_site_untransformed) # conver to matrix object
# convert to pairwise comparison
outer10_site_untransformed_comp <- outer10_site_untransformed_braymat[lower.tri(outer10_site_untransformed_braymat)] # extract lower triangle
outer10_site_untransformed_row_indices <- row(outer10_site_untransformed_braymat)[lower.tri(outer10_site_untransformed_braymat)] # extract row indices for first quad in comparison
outer10_site_untransformed_col_indices <- col(outer10_site_untransformed_braymat)[lower.tri(outer10_site_untransformed_braymat)] # extract col indices = second quad in comp
outer10_site_untransformed_loc1 <- rownames(outer10_site_untransformed_braymat)[outer10_site_untransformed_row_indices] # populate appropriate vector
outer10_site_untransformed_loc2 <- colnames(outer10_site_untransformed_braymat)[outer10_site_untransformed_col_indices] # populate appropriate vector
outer10_site_untransformed_pairwise_df <- data.frame(loc1 = outer10_site_untransformed_loc1, loc2 = outer10_site_untransformed_loc2, dissimilarity = outer10_site_untransformed_comp)
outer10_site_untransformed_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", outer10_site_untransformed_pairwise_df$loc1)) # extract timepoint from first quad in comp
outer10_site_untransformed_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", outer10_site_untransformed_pairwise_df$loc2)) # repeat for second quad
outer10_site_untransformed_pairwise_df$time <- abs(outer10_site_untransformed_pairwise_df$year2 - outer10_site_untransformed_pairwise_df$year1) # extract length of differences
outer10_site_untransformed_pairwise_df$site_1 <- sub(".*-(.*)", "\\1", outer10_site_untransformed_pairwise_df$loc1) # extract quadrat id from first quad in comp
outer10_site_untransformed_pairwise_df$site_2 <- sub(".*-(.*)", "\\1", outer10_site_untransformed_pairwise_df$loc2) # repeat for second quad

# outer 17

outer17_site_species_matrix <- as.matrix(outer17_site[ , -which(names(outer17_site) %in% meta_cols)])
outer17_site_species_matrix <- outer17_site_species_matrix/outer17_site$hard_cover # convert from percent to proportion of hard cover
rownames(outer17_site_species_matrix) <- paste(outer17_site$year,"-", outer17_site$site_habitat, sep = "")

# filter rare species
# outer17_site_species_presence <- colSums(outer17_site_species_matrix > 0) # Presence absence matrix
# outer17_site_species_percentage <- (outer17_site_species_presence/nrow(outer17_site_species_matrix)) # proportion of rows where a species is present
# outer17_site_common_species <- names(outer17_site_species_percentage[outer17_site_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
# outer17_site_species_matrix <- outer17_site_species_matrix[,colnames(outer17_site_species_matrix) %in% outer17_site_common_species] # filter original species matrix

outer17_site_rows_to_remove <- rowSums(outer17_site_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
outer17_site_species_matrix <- outer17_site_species_matrix[!outer17_site_rows_to_remove,]

outer17_site_species_matrix_untransformed <- outer17_site_species_matrix
outer17_site_species_matrix_transformed <- asin(sqrt(outer17_site_species_matrix)) # arcsin squareroot transform
outer17_site_subset <- outer17_site[!outer17_site_rows_to_remove,] # make sure rows are also removed from dataframe
outer17_site_subset$stratum <- paste(outer17_site_subset$year,"-", outer17_site_subset$site_habitat, sep = "") # for merging with bray matrix later

brayouter17_site <- vegdist(outer17_site_species_matrix_transformed, "bray") # create bray dissimilarity matrix
# View(brayouter17_site)
brayouter17_site_untransformed <- vegdist(outer17_site_species_matrix_untransformed, "bray")
outer17_site_braymat <- as.matrix(brayouter17_site) # conver to matrix object
# convert to pairwise comparison
outer17_site_comp <- outer17_site_braymat[lower.tri(outer17_site_braymat)] # extract lower triangle
outer17_site_row_indices <- row(outer17_site_braymat)[lower.tri(outer17_site_braymat)] # extract row indices for first quad in comparison
outer17_site_col_indices <- col(outer17_site_braymat)[lower.tri(outer17_site_braymat)] # extract col indices = second quad in comp
outer17_site_loc1 <- rownames(outer17_site_braymat)[outer17_site_row_indices] # populate appropriate vector
outer17_site_loc2 <- colnames(outer17_site_braymat)[outer17_site_col_indices] # populate appropriate vector
outer17_site_pairwise_df <- data.frame(loc1 = outer17_site_loc1, loc2 = outer17_site_loc2, dissimilarity = outer17_site_comp)
outer17_site_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", outer17_site_pairwise_df$loc1)) # extract timepoint from first quad in comp
outer17_site_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", outer17_site_pairwise_df$loc2)) # repeat for second quad
outer17_site_pairwise_df$time <- abs(outer17_site_pairwise_df$year2 - outer17_site_pairwise_df$year1) # extract length of differences
outer17_site_pairwise_df$site_1 <- sub(".*-(.*)", "\\1", outer17_site_pairwise_df$loc1) # extract quadrat id from first quad in comp
outer17_site_pairwise_df$site_2 <- sub(".*-(.*)", "\\1", outer17_site_pairwise_df$loc2) # repeat for second quad

outer17_site_untransformed_braymat <- as.matrix(brayouter17_site_untransformed) # conver to matrix object
# convert to pairwise comparison
outer17_site_untransformed_comp <- outer17_site_untransformed_braymat[lower.tri(outer17_site_untransformed_braymat)] # extract lower triangle
outer17_site_untransformed_row_indices <- row(outer17_site_untransformed_braymat)[lower.tri(outer17_site_untransformed_braymat)] # extract row indices for first quad in comparison
outer17_site_untransformed_col_indices <- col(outer17_site_untransformed_braymat)[lower.tri(outer17_site_untransformed_braymat)] # extract col indices = second quad in comp
outer17_site_untransformed_loc1 <- rownames(outer17_site_untransformed_braymat)[outer17_site_untransformed_row_indices] # populate appropriate vector
outer17_site_untransformed_loc2 <- colnames(outer17_site_untransformed_braymat)[outer17_site_untransformed_col_indices] # populate appropriate vector
outer17_site_untransformed_pairwise_df <- data.frame(loc1 = outer17_site_untransformed_loc1, loc2 = outer17_site_untransformed_loc2, dissimilarity = outer17_site_untransformed_comp)
outer17_site_untransformed_pairwise_df$year1 <- as.numeric(sub("(.*)-.*", "\\1", outer17_site_untransformed_pairwise_df$loc1)) # extract timepoint from first quad in comp
outer17_site_untransformed_pairwise_df$year2 <- as.numeric(sub("(.*)-.*", "\\1", outer17_site_untransformed_pairwise_df$loc2)) # repeat for second quad
outer17_site_untransformed_pairwise_df$time <- abs(outer17_site_untransformed_pairwise_df$year2 - outer17_site_untransformed_pairwise_df$year1) # extract length of differences
outer17_site_untransformed_pairwise_df$site_1 <- sub(".*-(.*)", "\\1", outer17_site_untransformed_pairwise_df$loc1) # extract quadrat id from first quad in comp
outer17_site_untransformed_pairwise_df$site_2 <- sub(".*-(.*)", "\\1", outer17_site_untransformed_pairwise_df$loc2) # repeat for second quad

@@

```{r site bray dissimilarity over time, fig.height=15, fig.width=15}
# for the purposes of this first set of analyses, I'm interested in dissimilarity of a given quad over time
# not necessarily interested in how quads compare to each other over time/at each timepoint
fringing_site_pairwise_subset <- fringing_site_pairwise_df[fringing_site_pairwise_df$site_1 == fringing_site_pairwise_df$site_2,]
backreef_site_pairwise_subset <- backreef_site_pairwise_df[backreef_site_pairwise_df$site_1 == backreef_site_pairwise_df$site_2,]
outer10_site_pairwise_subset <- outer10_site_pairwise_df[outer10_site_pairwise_df$site_1 == outer10_site_pairwise_df$site_2,]
outer17_site_pairwise_subset <- outer17_site_pairwise_df[outer17_site_pairwise_df$site_1 == outer17_site_pairwise_df$site_2,]

comb_site_subset <- rbind(fringing_site_pairwise_subset,
                     backreef_site_pairwise_subset,
                     outer10_site_pairwise_subset,
                     outer17_site_pairwise_subset)
# site_1 and site_2 columns are all the same now
# create new col just called for merge with other metadata
comb_site_subset$site_habitat <- comb_site_subset$site_1

bray_time_series_site <- left_join(unique(site_meta_df[ , c("site", "site_habitat", "habitat", "island_side")]), comb_site_subset, by = "site_habitat")

# transformed dissimilarities
bray_time_series_site$dissimilarity_trans <- sv_trans(bray_time_series_site$dissimilarity, bray_time_series_site)

bray_mod_site <- glmmTMB(dissimilarity_trans ~ time*habitat + (1 | site),
                              na.action = na.pass, family = beta_family(link = "logit"), data = bray_time_series_site) #
summary(bray_mod_site) 
hist(residuals(bray_mod_site)) # slight skew but not awful
plot(residuals(bray_mod_site) ~ fitted(bray_mod_site)) # some heteroskedasticity but not bad

emtrends(bray_mod_site, pairwise ~ habitat , "time") # no sig differences
bray_trends_emm_site <- emmip(bray_mod_site, habitat ~ time , at = list(time = seq(1,16,1)), plotit = F, CIs = T)

ggplot()+
   geom_point(data = bray_time_series_site, aes(x = time, y = dissimilarity, colour = habitat), alpha = 0.1 ) +
   geom_line(data = bray_trends_emm_site, aes(x = time, y = inv_logit(yvar), colour = habitat), linewidth = 2) +
   geom_ribbon(data = bray_trends_emm_site, aes(x = time, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = habitat), alpha = 0.3) +
   scale_colour_manual(values = habitat_colours) +
   scale_fill_manual(values = habitat_colours) +

   labs(y = "Dissimilarity\n") +
   theme_bw() + 
   theme(panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(),
         panel.background = element_blank(),
         plot.title = element_text(color = "grey20", size = 32, vjust = 0, face = "plain"),
         axis.line.y = element_line(colour = "black"),
         axis.text.y = element_text(color = "grey20", size = 28, hjust = .5, vjust = .5, face = "plain"),
         axis.title.y = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
         axis.line.x = element_line(colour = "black"),
         axis.text.x = element_text(color = "grey20", size = 28, angle = 90, vjust = 0.5, hjust=1, face = "plain"),
         axis.title.x = element_text(color = "grey20", size = 32, hjust = .5, vjust = 0, face = "plain"),
         strip.text.x = element_text(size = 32),
         legend.key.size = unit(2, "cm"), # increase legend size
         legend.text = element_text(size = rel(2)), # increase legend text
         legend.title = element_text(size = rel(2.25)),
         legend.position = 'bottom', # move legend to bottom
         legend.box = 'horizontal',
         legend.box.background = element_rect(colour = "black") )# make legend horizontal
```

```{r site macroalgae level bray curtis setup}
fringing_site_macro_species_matrix <- as.matrix(fringing_site_macro[ , -which(names(fringing_site_macro) %in% meta_cols)])
fringing_site_macro_species_matrix <- fringing_site_macro_species_matrix/fringing_site_macro$hard_cover # convert from percent to proportion of hard cover
rownames(fringing_site_macro_species_matrix) <- paste(fringing_site_macro$year,"-", fringing_site_macro$site_habitat, sep = "")

# filter rare species
# fringing_site_macro_species_presence <- colSums(fringing_site_macro_species_matrix > 0) # Presence absence matrix
# fringing_site_macro_species_percentage <- (fringing_site_macro_species_presence/nrow(fringing_site_macro_species_matrix)) # proportion of rows where a species is present
# fringing_site_macro_common_species <- names(fringing_site_macro_species_percentage[fringing_site_macro_species_percentage > 0.05]) # filter for presence in more than 5% of quads - only 9 species...
# fringing_site_macro_species_matrix <- fringing_site_macro_species_matrix[,colnames(fringing_site_macro_species_matrix) %in% fringing_site_macro_common_species] # filter original species matrix

fringing_site_macro_rows_to_remove <- rowSums(fringing_site_macro_species_matrix) == 0 # remove rows with no algal species observed - thought this was done already
fringing_site_macro_species_matrix <- fringing_site_macro_species_matrix[!fringing_site_macro_rows_to_remove,]

fringing_site_macro_species_matrix_untransformed <- fringing_site_macro_species_matrix
fringing_site_macro_species_matrix_transformed <- asin(sqrt(fringing_site_macro_species_matrix)) # arcsin squareroot transform
fringing_site_macro_subset <- fringing_site_macro[!fringing_site_macro_rows_to_remove,] # make sure rows are also removed from dataframe
fringing_site_macro_subset$stratum <- paste(fringing_site_macro_subset$year,"-", fringing_site_macro_subset$site_habitat, sep = "") # for merging with bray matrix later
